---
title: "Example models"
output:
  html_document:
    css: greta.css
    toc: yes
    toc_float:
      collapsed: false
    toc_depth: 3
    theme: lumen
    highlight: default
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Example models}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment = NA,
                      eval = greta:::check_tf_version("message"),
                      cache = TRUE)
library(greta)
```

## BUGS models

The BUGS project provide a number of example models written in the BUGS modelling language. These models will run in WinBUGS and OpenBUGS, and likely also in JAGS. The [Stan wiki](https://github.com/stan-dev/example-models/wiki/BUGS-Examples-Sorted-Alphabetically) provides Stan implementations of these models.

The following sections provide greta implementations of some of these example models, alongside the BUGS code from [WinBUGS examples volume 2](https://www.mrc-bsu.cam.ac.uk/wp-content/uploads/WinBUGS_Vol2.pdf) (pdf) and Stan code and an R version of the data from the [Stan example models wiki](https://github.com/stan-dev/example-models/wiki). 

### Air

*Air* analyses reported respiratory illness versus exposure to nitrogen dioxide in 103 children. The parameters `alpha`, `beta` and `sigma2` are known in advance, and the data are grouped into three categories.

See [WinBUGS examples volume 2](https://www.mrc-bsu.cam.ac.uk/wp-content/uploads/WinBUGS_Vol2.pdf) (pdf) for details.

#### data
<div class="data">
```{r air_data, highlight = FALSE}
y <- c(21, 20, 15)
n <- c(48, 34, 21)
Z <- c(10, 30, 50)
alpha <- 4.48
beta <- 0.76
sigma2 <- 81.14
sigma <- sqrt(sigma2)
tau <- 1 / sigma2
J <- 3
```
</div>

#### greta code
```{r air_greta}
theta <- normal(0, 32, dim = 2)
mu <- alpha + beta * Z
X <- normal(mu, sigma)
p <- ilogit(theta[1] + theta[2] * X)
distribution(y) <- binomial(n, p)
```

#### BUGS/JAGS code
<div class="bugs">
```
for(j in 1 : J) {
   y[j] ~ dbin(p[j], n[j])
   logit(p[j]) <- theta[1] + theta[2] * X[j]
   X[j] ~ dnorm(mu[j], tau)
   mu[j] <- alpha + beta * Z[j]
}
theta[1] ~ dnorm(0.0, 0.001)
theta[2] ~ dnorm(0.0, 0.001)
```
</div>

#### Stan code
<div class="stan">
```{r air_stan, echo = FALSE}
cat(readLines('https://raw.githubusercontent.com/stan-dev/example-models/master/bugs_examples/vol2/air/air.stan'), sep = '\n')
```
</div>

### Beetles

*Beetles* considers dose-response data from an experiment applying carbon disulphide to 8 beetles. The original example compares three different link functions; the logit, probit and complementary log-log. Here, only the code for the logit link is shown. You can implement the other two link functions in greta by changing `ilogit` to `iprobit` or `icloglog`.

See [WinBUGS examples volume 2](https://www.mrc-bsu.cam.ac.uk/wp-content/uploads/WinBUGS_Vol2.pdf) (pdf) for details.


#### data
<div class="data">
```{r beetles_data, highlight = FALSE}
x <- c(1.6907, 1.7242, 1.7552, 1.7842, 1.8113, 1.8369, 1.8610, 1.8839)
n <- c(59, 60, 62, 56, 63, 59, 62, 60)
r <- c(6, 13, 18, 28, 52, 53, 61, 60)
N <- 8
```
</div>

#### greta code
```{r beetles_greta}
alpha_star <- normal(0, 32)
beta <- normal(0, 32)
p <- ilogit(alpha_star + beta * (x - mean(x)))
distribution(r) <- binomial(n, p)

alpha <- alpha_star - beta * mean(x)
rhat <- p * n
```

#### BUGS/JAGS code
<div class="bugs">
```
for( i in 1 : N ) {
  r[i] ~ dbin(p[i],n[i])
  logit(p[i]) <- alpha.star + beta * (x[i] - mean(x[]))
  rhat[i] <- n[i] * p[i]
  culmative.r[i] <- culmative(r[i], r[i])
}
alpha <- alpha.star - beta * mean(x[])
beta ~ dnorm(0.0,0.001)
alpha.star ~ dnorm(0.0,0.001)
```
</div>

#### Stan code
<div class="stan">
```{r beetles_stan, echo = FALSE}
cat(readLines('https://raw.githubusercontent.com/stan-dev/example-models/master/bugs_examples/vol2/beetles/beetles_logit.stan'), sep = '\n')
```
</div>
