---
title: "Installing dependencies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installing dependencies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}

# Force limit to 2 cores for CRAN
Sys.setenv("R_PARALLELLY_AVAILABLE_CORES" = "2")
Sys.setenv("OMP_THREAD_LIMIT" = "2")
Sys.setenv("TF_NUM_INTRAOP_THREADS" = "1")
Sys.setenv("TF_NUM_INTEROP_THREADS" = "1")

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

```

```{r setup}
library(greta)
```

# Why greta needs Python

greta uses Google's [TensorFlow (TF)](https://www.tensorflow.org/) and
[TensorFlow Probability (TFP)](https://github.com/tensorflow/probability) under
the hood to do fast, scalable linear algebra and MCMC. TF and TFP are Python
packages, so greta needs a Python installation with those packages available.
This is different from how R package dependencies usually work, where CRAN builds
and manages everything for you.

The good news is that, thanks to recent improvements in reticulate, greta now
sets this up for you automatically. This vignette explains what happens by
default, how to choose a different Python environment, and how to install
dependencies yourself when you need to (for example, offline).

# Most users: nothing to install

By default greta installs its Python dependencies automatically using
[`uv`](https://docs.astral.sh/uv/) (via reticulate). The **first time you use greta in a session** -- for example when you create a greta array or use a
distribution -- reticulate downloads and sets up a compatible Python,
TensorFlow, and TensorFlow Probability for you.

So for most users, installation is just:

```{r first-use, eval = FALSE}
# if you haven't already
install.packages("greta")
library(greta)

# the first use triggers automatic setup of Python + TF + TFP
x <- normal(0, 1)
```

The first call may take a little while as the environment is installed; after
that it is cached and reused, so subsequent sessions start quickly.

You do **not** need to call `install_greta_deps()` for this to work --- that
function is for the conda / offline workflow described in
[Advanced: a managed conda environment](#advanced-a-managed-conda-environment).

# Checking what greta is using

`greta_sitrep()` reports the resolved Python backend and the versions in use.
It's the first thing to run if you're unsure what greta has picked up:

```{r sitrep, eval = FALSE}
greta_sitrep()
```

# Choosing a different Python environment

greta resolves which Python to use in this order:

1. The `RETICULATE_PYTHON` environment variable, which is usually set in
`~/.Renviron` or your shell environment.
2. Stored preference - set with `greta_set_python()`.
3. Auto-detected `greta-env-tf2` conda environment - created by `install_greta_deps()`.
4. The managed (uv) environment - the default, no setup needed. It's an
isolated Python environment that reticulate creates and manages for you
using the uv tool, so you don't need to install anything yourself.

If you have upgraded from an older greta that used a `greta-env-tf2` conda
environment (created by `install_greta_deps()`), greta detects and keeps using
it. So, upgrading to greta 0.6.0 does not change your setup.

To see which Python environment greta is using right now — and which it
will use after you restart R — run:

```{r show-sitrep, eval=FALSE}
greta_sitrep()
```

You can switch environments at any time with `greta_set_python()`, then
**restart R**. Each call reports what greta will resolve to on its next load:

```{r greta-set, eval = FALSE}
# use the managed (uv) environment (the default)
greta_set_python()
# use the "greta-env-tf2" conda environment
greta_set_python("conda")
# use a specific Python
greta_set_python("path", path = "/path/to/python")
# clear the choice; resolve automatically
greta_reset_python()
```

These store your choice (under `tools::R_user_dir("greta", "config")`) so it persists
across sessions.

For finer control, setting the `RETICULATE_PYTHON` environment variable to a
path (e.g. in your `.Rprofile`) takes precedence over the stored preference:

```{r reticulate-python, eval = FALSE}
Sys.setenv(RETICULATE_PYTHON = "/path/to/your/python")
library(greta)
```

Because `RETICULATE_PYTHON` takes precedence, a stored preference will
appear to be ignored while it is set. To go back to your stored
preference, remove `RETICULATE_PYTHON` from wherever it is set (for
example `~/.Renviron`, which you can open with
`usethis::edit_r_environ()`), then restart R.

# Advanced: a managed conda environment

The automatic (uv) setup covers most users. You might instead want to install
into a dedicated conda environment when you:

- are on an offline or restricted network (see [Using greta
  offline](#using-greta-offline)),
- need a specific, pinned, reproducible set of versions, or
- are setting up GPU / CUDA support.

`install_greta_deps()` builds a conda environment named `greta-env-tf2`,
installing TensorFlow and TensorFlow Probability into it. By default it uses TF
2.15.0, TFP 0.23.0, and Python 3.10:

```{r install-deps, eval = FALSE}
install_greta_deps()
```

Follow any prompts, then **restart R**. To make greta use this environment, set
it as your preference (then restart R again):

```{r use-conda, eval = FALSE}
greta_set_python("conda")
```

Using a conda environment isolates these exact Python modules from other Python
installations, so only greta sees them. This avoids a class of problems where,
for example, updating TensorFlow elsewhere on your machine would otherwise
overwrite the version greta needs.

## Choosing specific versions

`install_greta_deps()` takes a `deps` argument, built with `greta_deps_spec()`,
which lets you choose versions:

```{r deps-spec, eval = FALSE}
install_greta_deps(
  greta_deps_spec(
    tf_version = "2.15.0",
    tfp_version = "0.23.0",
    python_version = "3.10"
  )
)
```

If you specify versions of TF, TFP, and Python that are not compatible with each
other, `greta_deps_spec()` errors before installation begins and suggests
alternatives. The combinations greta knows about are recorded in the
`greta_deps_tf_tfp` dataset, which we built from
<https://www.tensorflow.org/install/source#tested_build_configurations>,
<https://www.tensorflow.org/install/source_windows#tested_build_configurations>,
and the TFP release notes. Inspect it with:

```{r deps-table, eval = FALSE}
View(greta_deps_tf_tfp)
```

To persist a version choice, store it with `greta_set_deps()`. The stored
versions are then used by both the managed (uv) environment (installed on the
next load) and `install_greta_deps()` (as its default `deps`):

```{r set-deps, eval = FALSE}
greta_set_deps(
  greta_deps_spec(
    tf_version = "2.14.0",
    tfp_version = "0.22.1",
    python_version = "3.10"
  )
)

# clear the stored choice and return to greta's defaults
greta_remove()
```

`install_greta_deps()` also takes `timeout` (minutes to wait before a component
times out, default 5) and `restart` (`"ask"` (default), `"force"`, or `"no"`).

## How `install_greta_deps()` works

For users who want the detail: greta runs the installation in a separate, clean R
session using [`callr`](https://callr.r-lib.org/), so Python and reticulate are
not already loaded. This also lets us route the large amount of console output
into a logfile, which you can open with `open_greta_install_log()`.

If miniconda isn't installed, greta installs it (a lightweight Python
distribution). If the `greta-env-tf2` environment doesn't exist, greta creates
it for a Python version compatible with the requested TF and TFP, then installs
the TF and TFP modules. In interactive RStudio sessions it then asks whether to
restart R.

# Using greta offline

Both the automatic (uv) setup and `install_greta_deps()` download packages from
the internet, so a blocked or air-gapped network will stop either approach. If
your network blocks PyPI but allows conda, the conda route via
`install_greta_deps()` may work where uv does not.

There are three escape routes, in order of robustness: point greta at a Python
environment that is already on disk, prime the managed (uv) environment once
while online, or pin dependency versions so an offline environment matches
what greta expects. `greta_sitrep()` reports whether your current setup is
ready to start offline.

## Point greta at an already-installed environment on disk

If you have a working Python with TF and TFP installed by some other means (for
example, provided by your institution, or copied onto an air-gapped machine),
point greta straight at it and restart R. These routes only ever *read* what is
already on disk — they never download anything — so they work with no network at
all.

`greta_set_python("path", ...)` accepts either a Python binary or an
environment directory (a virtualenv or conda prefix). Given a directory, greta
looks for `bin/python` inside it (`Scripts/python.exe` on Windows):

```{r offline, eval = FALSE}
# point at a specific Python binary
greta_set_python("path", path = "/opt/python-envs/greta/bin/python")

# or point at the environment directory and let greta find the binary
greta_set_python("path", path = "/opt/python-envs/greta")

# on Windows, an environment directory works the same way; greta looks for
# Scripts\\python.exe inside it
greta_set_python("path", path = "C:/python-envs/greta")
```

If your environment is a conda environment, you can instead select it by name
(this also never downloads anything):

```{r offline-conda, eval = FALSE}
greta_set_python("conda", name = "greta-env-tf2")
```

Alternatively, set the `RETICULATE_PYTHON` environment variable (for example in
`~/.Renviron`) to the Python binary, which takes precedence over any stored
preference.

## Prime the managed (uv) environment while online

If you use the default managed (uv) environment, greta starts offline once
reticulate's uv cache has been populated once (while online): run
`library(greta)` and use it once (for example `normal(0, 1)`), and you are
done. On its next start greta detects the populated cache and enables uv's
offline mode automatically, so it no longer reaches out to PyPI.

You can override this yourself by setting the `UV_OFFLINE` environment
variable before loading greta; greta never touches a value you have already
set:

```{r offline-override, eval = FALSE}
# force offline mode: e.g. before greta detects the cache, or on a machine
# where you copied the cache over
Sys.setenv(UV_OFFLINE = "1")

# force uv back online: for example to refresh the environment
Sys.setenv(UV_OFFLINE = "0")
```

To make either choice persistent, set `UV_OFFLINE=1` (or `UV_OFFLINE=0`) in
`~/.Renviron` instead (no `Sys.setenv()` needed there).

## Pin dependency versions

If someone else installs a Python environment for your offline machine, pin
the versions greta expects with `greta_set_deps()` so the two match (see
[Choosing specific versions](#choosing-specific-versions)). The stored
versions are used by both the managed (uv) environment and
`install_greta_deps()`.

## If the internet drops mid-workflow

If greta errors on first use with a message from uv like `error: Failed to
fetch` (or, with `UV_OFFLINE=1` set, `Network connectivity is disabled, but
the requested data wasn't found in the cache`), uv tried to resolve
dependencies without a working connection. To recover:

1. If greta's environment had already been installed on this machine, force
   offline mode with `Sys.setenv(UV_OFFLINE = "1")` (or in `~/.Renviron`),
   then restart R: uv then resolves from its cache without touching the
   network.
2. If it had not, reconnect once and use greta (for example `normal(0, 1)`)
   to prime the environment; after that greta starts offline automatically.
3. If you cannot reconnect, point greta at an already-installed Python with
   `greta_set_python("path", path = ...)` as above, and restart R.

# Troubleshooting

Installation doesn't always go to plan. Some things to try:

- **Restart R.** After changing the environment (with `install_greta_deps()`
  or `greta_set_python()`), you must restart R so greta can connect to the
  chosen Python.

- **Run `greta_sitrep()`.** It reports the resolved backend and the versions of
  Python, TF, and TFP, which often points straight at the problem.

- **Check the installation logfile.** If you used `install_greta_deps()`, open
  the logfile with `open_greta_install_log()` and search it (Ctrl/Cmd+F) for
  "error" or "warn". Even when there's no obvious fix, the logfile is useful to
  share on a forum or a [greta GitHub issue](https://github.com/greta-dev/greta/issues).

- **Start from a clean slate.** `reinstall_greta_deps()` removes miniconda and
  the greta conda environment and installs them again. To remove things yourself,
  use `greta_remove()`, which takes a `what` argument: `greta_remove("env")` for
  the conda environment, `greta_remove("miniconda")` for miniconda,
  `greta_remove("uv_cache")` for reticulate's uv cache, `greta_remove("preference")`
  for the stored Python preference, or `greta_remove()` (the default, `"all"`) for
  everything at once. Then reinstall with `install_greta_deps()`.

- **Check internet access.** Installing dependencies needs a connection, and some
  networks block package downloads. See [Using greta
  offline](#using-greta-offline).

If the helpers don't work, you can install the Python modules yourself into a
conda environment:

```{r manual-conda, eval = FALSE}
reticulate::install_miniconda()
reticulate::conda_create(
  envname = "greta-env-tf2",
  python_version = "3.10"
)
reticulate::conda_install(
  envname = "greta-env-tf2",
  packages = c(
    "tensorflow-probability==0.23.0",
    "tensorflow==2.15.0"
  )
)
```

Then point greta at it with `greta_set_python("conda")` and restart R.
