### Multiple species with sub-model for regression coefficients

*Multiple species with sub-model for regression coefficients* This is an example of a logistic regression being used to estimate the probability of multiple species' presences along a number of environmental gradients. Instead of assuming independence of species regression coefficients, or partial pooling in shared distributions, we use a sub-model to estimate species regression coefficients. In this case, we're using species traits to estimate their response to different environmental gradients.

Because we're building a sub-model, it's more efficient to simply add a coloumn of ones to dataframes for the base model and sub-model. This is simply to prevent our code fmor becoming too cumbersome. If we didnt want to use our sub-model to estimate the intercept, we would not need to include the coloum of ones in the environemntal dataframe.

We first simulate some data to model followed by the `greta` code.

#### data
<div class="data">
```{r data_multilevel, highlight = FALSE}
# make fake data
n_species <- 3
n_env <- 1
n_sites <- 5
n_traits <- 1

env <- matrix(rnorm(n_sites * n_env), nrow = n_sites) # create n_sites x n_species matrix of environmental variables
traits <- matrix(rnorm(n_species * n_traits), nrow = n_species) # create n_species * n_traits matix of trait variables
occupancy <- matrix(rbinom(n_sites * n_species, 1, 0.5), nrow = n_species) # create n_sites * n_species matrix of observed occupancy
```
</div>

#### greta code
```{r multilevel_greta}
# load greta
library(greta)

# data wrangling
traits <- cbind(rep(1, n_species), traits) # include a coloum of 1's for sub-model intercept estimation
env <- cbind(rep(1, n_sites), env) # include a coloum of 1's for base model intercept estimation

# redefine the n_env and n_traits after adding in coloum of 1's for intercepts
n_env <- ncol(env)
n_traits <- ncol(traits)

# create greta dataframes from data matrices
X <- as_data(env)
Y <- as_data(occupancy)
U <- as_data(traits)

# create empty greta dataframes for variables to be estimated
g <- normal(0, 10, dim = c(n_traits, n_env)) # sub-model parameters are drawn from normal distributions
beta <- U %*% t(g) # parameters of the base model are a function of the parameters of the sub-model

env_effect <- beta %*% t(X) # transpose environmental matrix X for matrix multiplication

linear_predictor <- env_effect # In this case, env_effect already includes the intercept and so it is equal to the final linear predictor
# inverse logit of linear predictor
p <- ilogit(linear_predictor)

# data are bernoulli distributed
distribution(Y) <- bernoulli(p)

# create model object with the parameters we want samples of
m <- model(beta, g)

# plot model structure
plot(m)
```
