---
output: html_document
---
### Multiple species modelling independently and concurrently

*Multiple species modelled independently and concurrently* This is an example of a logistic regression being used to estimate the probability of multiple species' presences along a number of environmental gradients. Although modelled concorrently, the random variables for each species are independent. We first simulate some data to model followed by the `greta` code.

Where a single observation per species and location would have a bernoulli error distribution, multiple observations for each species and location have a binomial distribution. The small change in the code is commented below.

When modelling multiple species (or other grouping factor), we need an extra step in constructing the linear predictor. In order to add multiple `greta` arrays together *for each species* we can use the `sweep()` function.

#### data
<div class="data">
```{r data_multispecies_bernoulli, highlight = FALSE}
# make fake data
n_species <- 5
n_env <- 3
n_sites <- 20

env <- matrix(rnorm(n_sites * n_env), nrow = n_sites)
occupancy <- matrix(rbinom(n_species * n_sites, 1, 0.5), nrow = n_sites)
```
</div>

#### greta code
```{r multispecies_bernoulli_greta}
# load greta
library(greta)

# data
X <- as_data(env)
Y <- as_data(occupancy)

# variables
alpha <- normal(0,10, dim = n_species)
beta <- normal(0, 10, dim = c(n_env, n_species))

env_effect <- X %*% beta

# matrix addition with `sweep()` create interim variable
linear_predictor <- sweep(env_effect, 2, alpha, FUN = '+')

# ilogit of linear predictor
p <- ilogit(linear_predictor)

# a single observation means our data are bernoulli distributed
distribution(Y) <- bernoulli(p)
# distribution(Y) <- binomial(n_obs, p) # for n repeat observations per species/ location combination, we can use the binomial error distribution

# create model object with the things we want samples of
m <- model(alpha, beta)
```
