---
output: html_document
---
### Multiple species modelling independently and concurrently

*Multiple species modelled independently and concurrently* This is an example of a simple poisson regression being used to estimate the abundance of a species along a number of environmental gradients. We first simulate some data to model followed by the `greta` code.

When modelling multiple species (or other grouping factor), we need an extra step in constructing the linear predictor. In order to add multiple `greta` arrays together *for multiple species* we can use the `sweep()` function.

#### data
<div class="data">
```{r data, highlight = FALSE}
# make fake data
n_species <- 5
n_env <- 3
n_sites <- 20

env <- matrix(rnorm(n_sites * n_env), nrow = n_sites)
occupancy <- matrix(rbinom(n_species * n_sites, 1, 0.5), nrow = n_species)
```
</div>

#### greta code
```{r multispecies_bernoulli_greta}
# load greta
library(greta)

# data
X <- as_data(t(env))
Y <- as_data(occupancy)

# variables
alpha <- normal(0,10, dim = c(n_species, 1))
beta <- normal(0, 10, dim = c(n_species, n_env))

env_effect <- beta %*% X #not transposed environmental matrix

# matrix addition with `sweep()` create interim variable
linear_predictor <- sweep(env_effect, 1, alpha, FUN = '+')

# ilogit of linear predictor
p <- ilogit(linear_predictor)

# a single observation means our data are bernoulli distributed
distribution(Y) <- bernoulli(p)

# create model object with the things we want samples of
m <- model(alpha, beta)

# plot model structure
plot(m)
```
